A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-09-04 papers/water-saft/paper.tex
Small changes to water paper Tested everything on node38.

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT)~\cite{chapman1989saft}.  SAFT is a theory based on a model of
hard spheres with weak dispersion interactions and hydrogen-bonding
association sites, which has been used to accurately model the
equations of state of both pure fluids and mixtures over a wide range
of temperatures and pressures~\cite{muller2001molecular,
  tan2008recent}.  The association term in SAFT uses Wertheim's
first-order thermodynamic perturbation theory to describe an
associating fluid as hard-spheres with strong associative interactions
at specific sites on the surface of each sphere~\cite{wertheim1984fluidsI,
  wertheim1984fluidsII, wertheim1986fluidsIII, wertheim1986fluidsIV}.
These association sites have an attractive interaction at contact, and
rely on the hard-sphere pair distribution function at contact
$g_\sigma^\text{HS}$ in order to determine the extent of association.
While this function is known for the homogeneous hard-sphere
fluid, it must be approximated for inhomogeneous systems, such as occur
at liquid interfaces.

@@

As in \hughesetal, we use Clark's five empirical parameters, and fit
the calculated surface tension to experimental surface tension at
ambient conditions by tuning the parameter $\lscale$, which adjusts
the length-scale of the average density used for the dispersion
interaction.  With the improved association term, we find these agree
when $\lscale$ is 0.454, which is an increase from the value of 0.353
found by \hughesetal.  In order to explore further the change made by
the improved association term, we compared the new functional with
that of \hughesetal\ for the two hydrophobic cases of hard rods and
the hard spherical solutes.

@@

In addition to the density, we examine the number of hydrogen bonds
which are broken due to the presence of a hard rod.  We define this
quantity as
\begin{align}
  N_{\text{broken HB}} &= 2 \int (X(\rr) - X_{\text{bulk}})n_{\text{site}}(\rr) d\rr
\end{align}
where $X_{\text{bulk}} = 0.1?$ \fixme{get this number} is the fraction
of unbonded association sites in the bulk.  The factor of 2 is chosen
to account for the four association sites per molecule, and the fact
that each broken hydrogen bond must be represented twice---once for
each of the molecules involved.  In
Fig.~\ref{fig:single-rod-broken-HB} we show the number of hydrogen
bonds broken by a hard rod per nanometer length, as predicted by the
functional of \hughesetal\ (dashed line) and this work (solid line),
as a function of the radius of the hard rod.  In each case in the
limit of large rods, the number of broken bonds is proportional to the
surface area.  At every radius, the functional of
\hughesetal\ predicts approximately four times as many broken hydrogen
bonds as the improved functional.

>>>>

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT)~\cite{chapman1989saft}.  SAFT is a theory based on a model of
hard spheres with weak dispersion interactions and hydrogen-bonding
association sites, which has been used to accurately model the
equations of state of both pure fluids and mixtures over a wide range
of temperatures and pressures~\cite{muller2001molecular,
  tan2008recent}. The association contribution to the free energy uses
Wertheim's first-order thermodynamic perturbation theory to describe
an associating fluid as hard-spheres with strong associative
interactions at specific sites on the surface of each
sphere~\cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV}.  These association
sites have an attractive interaction at contact, and rely on the
hard-sphere pair distribution function at contact $g_\sigma^\text{HS}$
in order to determine the extent of association.  While this function
is known for the homogeneous hard-sphere fluid, it must be
approximated for inhomogeneous systems, such as occur at liquid
interfaces.

@@

As in \hughesetal, we use Clark's five empirical parameters, and fit
the calculated surface tension to experimental surface tension at
ambient conditions by tuning the parameter $\lscale$, which adjusts
the length-scale of the average density used for the dispersion
interaction.  With the improved association term, we find these agree
when $\lscale$ is 0.454, which is an increase from the value of 0.353
found by \hughesetal.  In order to explore further the change made by
the improved association term, we compared the new functional with
that of \hughesetal\ for the two hydrophobic cases of the hard rod and
the hard spherical solute.

@@

In addition to the density, we examine the number of hydrogen bonds
which are broken due to the presence of a hard rod.  We define this
quantity as
\begin{align}
  N_{\text{broken HB}} &= 2 \int (X(\rr) - X_{\text{bulk}})n_{\text{site}}(\rr) d\rr
\end{align}
where $X_{\text{bulk}} = 0.13$ is the fraction of unbonded
association sites in the bulk.  The factor of 2 is chosen to account
for the four association sites per molecule, and the fact that each
broken hydrogen bond must be represented twice---once for each of the
molecules involved.  In Fig.~\ref{fig:single-rod-broken-HB} we show
the number of hydrogen bonds broken by a hard rod per nanometer
length, as predicted by the functional of \hughesetal\ (dashed line)
and this work (solid line), as a function of the radius of the hard
rod.  In each case in the limit of large rods, the number of broken
bonds is proportional to the surface area.  At every radius, the
functional of \hughesetal\ predicts approximately four times as many
broken hydrogen bonds as the improved functional.
